---
title: "Diabetes_Example"
author: "Jonas Mikhaeil"
date: "2024-01-29"
output: html_document
---
```{r,message=False}
library(tidyverse)
library(readr)
library(janitor)
library("rstan")
library(patchwork)
library(grid)
library("caret")
library(patchwork)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```


## Read in data

NHANES data provided by  https://github.com/madisoncoots/race-in-diabetes-screening/tree/main (Coots, 2023).
We normalize the weights to have mean 1.

```{r}
data <- readRDS(file = "data/diabetes/processed/data.rds")
data$wtmec8yr <- data$wtmec8yr / mean(data$wtmec8yr)

formula1 <-
  diabetes_diagnosis ~ health_insurance + gender + ridageyr + bmxbmi + (ridageyr *
                                                                          bmxbmi) +
  whd140 + bmxwt + bmxht + bmxwaist + relatives_had_diabetes + felt_depressed +
  income + as.factor(ntile(data$bmxbmi, 10)) + food_security

formula2 <-
  ~ diabetes + diabetes_diagnosis + wtmec8yr + gender + ridageyr + bmxbmi + (ridageyr *
                                                                               bmxbmi) +
  whd140 + bmxwt + bmxht + bmxwaist + relatives_had_diabetes + felt_depressed +
  income + health_insurance  + food_security
### create data-set by selecting all relevant variables and cleaned of NaNs
data_clean <- data %>%
  dplyr::select(all.vars(formula2)) %>%
  na.omit()
###create model matrix for stan
data <- data_clean
X <- model.matrix(formula1 , data = data)
```

### Fit classical logistic regression models
Fit simple, complex and ground-truth model

```{r}
simple_formula <-
  diabetes_diagnosis ~ gender + ridageyr + bmxbmi + (ridageyr * bmxbmi) + as.factor(ntile(data$bmxbmi, 10)) +
  whd140 + bmxwt + bmxht + bmxwaist + relatives_had_diabetes + felt_depressed +
  income + food_security
simple_model <- glm(
  simple_formula,
  data = data,
  family = quasibinomial("logit"),
  weights = wtmec8yr
)
simple_model_pred <-
  predict(simple_model, newdata = data, type = "response")

complex_formula <-
  diabetes_diagnosis ~ health_insurance + gender + ridageyr + bmxbmi + (ridageyr *
                                                                          bmxbmi) + as.factor(ntile(data$bmxbmi, 10)) +
  whd140 + bmxwt + bmxht + bmxwaist + relatives_had_diabetes + felt_depressed +
  income + food_security
complex_model <- glm(
  complex_formula,
  data = data,
  family = quasibinomial("logit"),
  weights = wtmec8yr
)
complex_model_pred <-
  predict(complex_model, newdata = data, type = "response")

#Full model to approximate ground-truth diabetes incidence rates
oracle_model_formula <-
  diabetes ~ gender + ridageyr + bmxbmi + (ridageyr * bmxbmi) + as.factor(ntile(data$bmxbmi, 10)) +
  whd140 + bmxwt + bmxht + bmxwaist + relatives_had_diabetes + felt_depressed +
  income + health_insurance  + food_security

oracle_model <- glm(
  oracle_model_formula,
  data = data,
  family = quasibinomial("logit"),
  weights = wtmec8yr
)

oracle_model_pred <-
  predict(oracle_model, newdata = data, type = "response")
```



## Measurement Model

### Determining thresholds from prior knowledge
We know the total rate of diabetes in the population is around $14\%$, choose mean accordingly.

Base-rate should be roughly 14%.
```{r}
alpha <- -1.8
x <- rlogis(100000, alpha, 1)
mean(x > 0)
```

Determine thresholds based on base-rate:
Uninsured: roughly 29% undiagnosed
```{r}
beta1 <- -0.38
y <- rlogis(100000, alpha + beta1, 1)
1 - mean(y > 0) / mean(x > 0)
```
Insured: roughly 16% undiagnosed
```{r}
beta2 <- -0.22
y <- rlogis(100000, alpha + beta2, 1)
1 - mean(y > 0) / mean(x > 0)
```

### Fit MLE version of measurement model 

Maximum-likelihood version of measurement model (does not incorporate term $e$ for idiosyncrasies).
```{r}
errorModel_formula <-
  diabetes_diagnosis ~ gender + ridageyr + bmxbmi + (ridageyr * bmxbmi) + as.factor(ntile(data$bmxbmi, 10)) +
  whd140 + bmxwt + bmxht + bmxwaist + relatives_had_diabetes + felt_depressed +
  income + health_insurance  + food_security
errorModel <- glm(
  errorModel_formula,
  data = data,
  family = quasibinomial("logit"),
  weights = wtmec8yr
)

EM_predict <- function(model, data, shift) {
  alpha <- errorModel$coefficients["(Intercept)"]
  beta <- errorModel$coefficients["health_insuranceYes"]
  alpha <- alpha + shift[1]
  beta <- beta + shift[2]
  errorModel$coefficients["(Intercept)"] <- alpha
  errorModel$coefficients["health_insuranceYes"] <- beta
  predict(errorModel, newdata = data, type = "response")
}

errorModel_pred <- EM_predict(errorModel, data, c(0.38, -0.17))  
```

### Threshold model with Stan

Here we run the following Hierarchical Bayesian measurement model, where the thresholds are as determined above.
$$
\begin{align} \nonumber
y \, &= \, \begin{cases}
    1 \, \, \text{if}\, \, u_2 \geq 0 \\
    0 \, \, \text{else}
\end{cases} \\[2pt] \nonumber
    u_1 | \beta \, &\sim \, \text{logistic}(X\beta,1) \\[2pt] 
    u_2 \, &= \,  u_1 - t(X) - e \\[2pt] \nonumber
    u_3 \, &= \, \begin{cases}
    1 \, \, \text{if}\, \, u_1 \geq 0 
    \\[2pt] \nonumber
    0 \, \, \text{else}
\end{cases} \\[2pt] \nonumber
e \, &\sim \, \text{normal}^+(0,0.1).
\end{align}
$$
The model is implemented in logistic_regression_diabetes.rstan.
Running the model takes some time to run. Predictions from an inferred model are saved as "diabetes_thresholdModel_pred" to ease reproduction of our plots.
```{r}
data_stan <- list(
  N = dim(X)[1],
  K = dim(X)[2],
  diagnosis = data$diabetes_diagnosis,
  weights = data$wtmec8yr,
  x = X,
  shift1 = 0.38,
  shift2 = -0.17
)
init_fun_log <- function() {
  list(beta = rnorm(dim(X)[2], 0, .1))
}
fit_bayes_stan <-
  stan(
    file = "logistic_regression_diabetes.rstan",
    data = data_stan,
    iter = 2000,
    chains = 4,
    init = init_fun_log,
    refresh = 1000
  )
#saveRDS(fit_bayes_stan,file="measurementModel_fullX")
```

```{r}
#fit_bayes_stan<-readRDS("measurementModel_fullX")
#measurement_model_pred <- summary(fit_bayes_stan,pars="pred")$summary[ , "50%"]
#saveRDS(measurement_model_pred,"diabetes_thresholdModel_pred")
measurement_model_pred <- readRDS("diabetes_thresholdModel_pred")
```


## Plotting

### Calibration plot: Comparison with empirical rate determined with logistic regression on true labels

```{r}
risk_score_upper_bound <- 0.3
incidence_upper_bound <- 0.3
scale_factor <- 1 / 0.01

simple_calibration_plot_data <- data %>%
  mutate(risk_score = simple_model_pred,
         est_diabetes_prob = oracle_model_pred) %>%
  filter(!is.na(risk_score), !is.na(est_diabetes_prob)) %>%
  dplyr::select(health_insurance,
                risk_score,
                est_diabetes_prob,
                wtmec8yr,
                diabetes) %>%
  mutate(risk_score_bin = round((risk_score) * scale_factor) / scale_factor) %>% # round to scale factor
  group_by(health_insurance, risk_score_bin) %>%
  summarize(
    n_in_bin = sum(wtmec8yr),
    bin_avg_risk_score = sum(risk_score * wtmec8yr) / sum(wtmec8yr),
    diabetes_prev = sum(est_diabetes_prob * wtmec8yr) / sum(wtmec8yr),
    ,
    diabetes_prev_emp = sum(diabetes * wtmec8yr) / sum(wtmec8yr)
  )


complex_calibration_plot_data <- data %>%
  mutate(risk_score = complex_model_pred,
         est_diabetes_prob = oracle_model_pred) %>%
  filter(!is.na(risk_score), !is.na(est_diabetes_prob)) %>%
  dplyr::select(health_insurance,
                risk_score,
                est_diabetes_prob,
                wtmec8yr,
                diabetes) %>%
  mutate(risk_score_bin = round((risk_score) * scale_factor) / scale_factor) %>%
  group_by(health_insurance, risk_score_bin) %>%
  summarize(
    n_in_bin = sum(wtmec8yr),
    bin_avg_risk_score = sum(risk_score * wtmec8yr) / sum(wtmec8yr),
    diabetes_prev = sum(est_diabetes_prob * wtmec8yr) / sum(wtmec8yr),
    diabetes_prev_emp = sum(diabetes * wtmec8yr) / sum(wtmec8yr)
  )



plot1 <- simple_calibration_plot_data %>%
  ggplot(aes(
    x = bin_avg_risk_score,
    y = diabetes_prev,
    color = as.factor(health_insurance)
  )) +
  scale_color_manual(values = c("#F88379", "#A52A2A")) +
  geom_line() +
  geom_point() +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "darkgray"
  ) +
  xlab("Predicted risk") +
  ylab("Observed diabetes rate") +
  scale_y_continuous(
    labels = scales::percent,
    expand = c(0, 0),
    breaks = seq(0.0, incidence_upper_bound, 0.05)
  ) +
  scale_x_continuous(
    labels = scales::percent,
    expand = c(0, 0.01),
    breaks = seq(0.0, risk_score_upper_bound, 0.05)
  ) +
  coord_cartesian(
    xlim = c(0, risk_score_upper_bound),
    ylim = c(0, incidence_upper_bound)
  ) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "none",
    axis.title.x = element_blank()
  )


plot2 <- complex_calibration_plot_data %>%
  ggplot(aes(
    x = bin_avg_risk_score,
    y = diabetes_prev,
    color = as.factor(health_insurance)
  )) +
  scale_color_manual(values = c("#7393B3", "#0F52BA")) +
  geom_line() +
  geom_point() +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "darkgray"
  ) +
  xlab("Predicted risk") +
  ylab("Observed diabetes rate") +
  annotate(
    geom = "text",
    x = 0.15,
    y = 0.03,
    label = "health insurance",
    col = "#0F52BA",
    size = 3.5
  ) +
  annotate(
    geom = "text",
    x = 0.1,
    y = 0.29,
    label = "no health insurance",
    col = "#7393B3",
    size = 3.5
  ) +
  
  scale_y_continuous(
    labels = scales::percent,
    expand = c(0, 0),
    breaks = seq(0.0, incidence_upper_bound, 0.05)
  ) +
  scale_x_continuous(
    labels = scales::percent,
    expand = c(0, 0.01),
    breaks = seq(0.0, risk_score_upper_bound, 0.05)
  ) +
  coord_cartesian(
    xlim = c(0, risk_score_upper_bound),
    ylim = c(0, incidence_upper_bound)
  ) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "none",
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
  )
```







```{r}
EM_calibration_plot_data <- data %>%
  mutate(risk_score = measurement_model_pred,
         est_diabetes_prob = oracle_model_pred) %>%
  filter(!is.na(risk_score),!is.na(est_diabetes_prob)) %>%
  dplyr::select(health_insurance,
                risk_score,
                est_diabetes_prob,
                wtmec8yr,
                diabetes) %>%
  mutate(risk_score_bin = round((risk_score) * scale_factor) / scale_factor) %>% # round to the nearest 0.005
  group_by(health_insurance, risk_score_bin) %>%
  summarize(
    n_in_bin = sum(wtmec8yr),
    bin_avg_risk_score = sum(risk_score * wtmec8yr) / sum(wtmec8yr),
    diabetes_prev = sum(est_diabetes_prob * wtmec8yr) / sum(wtmec8yr),
    diabetes_prev_emp = sum(diabetes * wtmec8yr) / sum(wtmec8yr)
  )

plot3 <- EM_calibration_plot_data %>%
  ggplot(aes(
    x = bin_avg_risk_score,
    y = diabetes_prev,
    color = as.factor(health_insurance)
  )) +
  scale_color_manual(values = c("#ffc87c", "#ff8c00")) +
  geom_line() +
  geom_point() +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "darkgray"
  ) +
  xlab("Predicted risk") +
  ylab("Observed diabetes rate") +
  scale_y_continuous(
    labels = scales::percent,
    expand = c(0, 0),
    breaks = seq(0.0, incidence_upper_bound, 0.05)
  ) +
  scale_x_continuous(
    labels = scales::percent,
    expand = c(0, 0),
    breaks = seq(0.0, risk_score_upper_bound, 0.05)
  ) +
  coord_cartesian(
    xlim = c(0, risk_score_upper_bound),
    ylim = c(0, incidence_upper_bound)
  ) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "none",
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  )

```




```{r}
#pdf("diabetes_riskPred.pdf",9,3)
layout <- plot1 + plot2 + plot3
wrap_elements(layout) +
  theme(plot.margin = margin(
    t = 0,
    r = 0,
    b = 0,
    l = 0,
    unit = "pt"
  )) 
```
### Calibration plot: Comparison with empirical rate determined by naive binning - very noisy.


```{r}
risk_score_upper_bound <- 0.2
incidence_upper_bound <- 0.2
scale_factor <- 1 / 0.05
simple_calibration_plot_data <- data %>%
  mutate(risk_score = simple_model_pred,
         est_diabetes_prob = oracle_model_pred) %>%
  filter(!is.na(risk_score),!is.na(est_diabetes_prob)) %>%
  dplyr::select(health_insurance,
                risk_score,
                est_diabetes_prob,
                wtmec8yr,
                diabetes) %>%
  mutate(risk_score_bin = round((risk_score) * scale_factor) / scale_factor) %>% # round to the nearest 0.005
  group_by(health_insurance, risk_score_bin) %>%
  summarize(
    n_in_bin = sum(wtmec8yr),
    bin_avg_risk_score = sum(risk_score * wtmec8yr) / sum(wtmec8yr),
    diabetes_prev = sum(est_diabetes_prob * wtmec8yr) / sum(wtmec8yr),
    ,
    diabetes_prev_emp = sum(diabetes * wtmec8yr) / sum(wtmec8yr)
  )


complex_calibration_plot_data <- data %>%
  mutate(risk_score = complex_model_pred,
         est_diabetes_prob = oracle_model_pred) %>%
  filter(!is.na(risk_score),!is.na(est_diabetes_prob)) %>%
  dplyr::select(health_insurance,
                risk_score,
                est_diabetes_prob,
                wtmec8yr,
                diabetes) %>%
  mutate(risk_score_bin = round((risk_score) * scale_factor) / scale_factor) %>% # round to the nearest 0.005
  group_by(health_insurance, risk_score_bin) %>%
  summarize(
    n_in_bin = sum(wtmec8yr),
    bin_avg_risk_score = sum(risk_score * wtmec8yr) / sum(wtmec8yr),
    diabetes_prev = sum(est_diabetes_prob * wtmec8yr) / sum(wtmec8yr),
    diabetes_prev_emp = sum(diabetes * wtmec8yr) / sum(wtmec8yr)
  )



plot1_emp <- simple_calibration_plot_data %>%
  ggplot(aes(
    x = bin_avg_risk_score,
    y = diabetes_prev_emp,
    color = as.factor(health_insurance)
  )) +
  scale_color_manual(values = c("#F88379", "#A52A2A")) + #D22B2B
  geom_line() +
  geom_point() +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "darkgray"
  ) +
  xlab("Predicted risk") +
  ylab("Observed diabetes rate") +
  scale_y_continuous(
    labels = scales::percent,
    expand = c(0, 0),
    breaks = seq(0.0, incidence_upper_bound, 0.05)
  ) +
  scale_x_continuous(
    labels = scales::percent,
    expand = c(0, 0.01),
    breaks = seq(0.0, risk_score_upper_bound, 0.05)
  ) +
  coord_cartesian(
    xlim = c(0, risk_score_upper_bound),
    ylim = c(0, incidence_upper_bound)
  ) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "none",
    axis.title.x = element_blank()
  )


plot2_emp <- complex_calibration_plot_data %>%
  ggplot(aes(
    x = bin_avg_risk_score,
    y = diabetes_prev_emp,
    color = as.factor(health_insurance)
  )) +
  scale_color_manual(values = c("#7393B3", "#0F52BA")) +
  geom_line() +
  geom_point() +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "darkgray"
  ) +
  xlab("Predicted risk") +
  ylab("Observed diabetes rate") +
  annotate(
    geom = "text",
    x = 0.1,
    y = 0.03,
    label = "health insurance",
    col = "#0F52BA",
    size = 3.5
  ) +
  annotate(
    geom = "text",
    x = 0.06,
    y = 0.19,
    label = "no health insurance",
    col = "#7393B3",
    size = 3.5
  ) +
  
  
  scale_y_continuous(
    labels = scales::percent,
    expand = c(0, 0),
    breaks = seq(0.0, incidence_upper_bound, 0.05)
  ) +
  scale_x_continuous(
    labels = scales::percent,
    expand = c(0, 0.01),
    breaks = seq(0.0, risk_score_upper_bound, 0.05)
  ) +
  coord_cartesian(
    xlim = c(0, risk_score_upper_bound),
    ylim = c(0, incidence_upper_bound)
  ) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "none",
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  )
```


```{r}
EM_calibration_plot_data <- data_clean %>%
  mutate(risk_score = measurement_model_pred,
         est_diabetes_prob = oracle_model_pred) %>%
  filter(!is.na(risk_score),!is.na(est_diabetes_prob)) %>%
  dplyr::select(health_insurance,
                risk_score,
                est_diabetes_prob,
                wtmec8yr,
                diabetes) %>%
  mutate(risk_score_bin = round((risk_score) * scale_factor) / scale_factor) %>% # round to the nearest 0.005
  group_by(health_insurance, risk_score_bin) %>%
  summarize(
    n_in_bin = sum(wtmec8yr),
    bin_avg_risk_score = sum(risk_score * wtmec8yr) / sum(wtmec8yr),
    diabetes_prev = sum(est_diabetes_prob * wtmec8yr) / sum(wtmec8yr),
    diabetes_prev_emp = sum(diabetes * wtmec8yr) / sum(wtmec8yr)
  )
plot3_emp <- EM_calibration_plot_data %>%
  ggplot(aes(
    x = bin_avg_risk_score,
    y = diabetes_prev_emp,
    color = as.factor(health_insurance)
  )) +
  scale_color_manual(values = c("#ffc87c", "#ff8c00")) +
  geom_line() +
  geom_point() +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "darkgray"
  ) +
  xlab("Predicted risk") +
  ylab("Observed diabetes rate") +
  
  scale_y_continuous(
    labels = scales::percent,
    expand = c(0, 0),
    breaks = seq(0.0, incidence_upper_bound, 0.05)
  ) +
  scale_x_continuous(
    labels = scales::percent,
    expand = c(0, 0.01),
    breaks = seq(0.0, risk_score_upper_bound, 0.05)
  ) +
  coord_cartesian(
    xlim = c(0, risk_score_upper_bound),
    ylim = c(0, incidence_upper_bound)
  ) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "none",
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  )
```


```{r}
pdf("diabetes_riskPred_emp.pdf",9,3)
layout <- plot1_emp + plot2_emp + plot3_emp
wrap_elements(layout) +
  theme(plot.margin = margin(
    t = 0,
    r = 0,
    b = 0,
    l = 0,
    unit = "pt"
  )) 
```

## Measures of Accuracy

```{r}
log_score <- function(data, pred) {
  mean(data * log(pred) + (1 - data) * log(1 - pred))
}
brier_score <- function(data, pred) {
  mean(data * (2 * pred - (pred ^ 2 + (1 - pred) ^ 2) - 1) + (1 - data) *
         (2 * (1 - pred) - (pred ^ 2 + (1 - pred) ^ 2) - 1))
}

MSE <- function(data, pred) {
  mean((round(pred) - data)) ^ 2
  
}
accuracy <-  function(data, pred) {
  length(pred[(round(pred) - data) == 0]) / length(data)
}

log_score(data$diabetes, simple_model_pred)
log_score(data$diabetes, complex_model_pred)
log_score(data$diabetes, measurement_model_pred)
log_score(data$diabetes, oracle_model_pred)
brier_score(data$diabetes, simple_model_pred)
brier_score(data$diabetes, complex_model_pred)
brier_score(data$diabetes, measurement_model_pred)
brier_score(data$diabetes, oracle_model_pred)
MSE(data$diabetes, simple_model_pred)
MSE(data$diabetes, complex_model_pred)
MSE(data$diabetes, measurement_model_pred)
MSE(data$diabetes, oracle_model_pred)

accuracy(data$diabetes, simple_model_pred)
accuracy(data$diabetes, complex_model_pred)
accuracy(data$diabetes, measurement_model_pred)
accuracy(data$diabetes, oracle_model_pred)
confusionMatrix(
  as.factor(round(simple_model_pred)),
  as.factor(round(data$diabetes)),
  positive = "1",
  dnn = c("Prediction", "Reference"),
)
confusionMatrix(
  as.factor(round(complex_model_pred)),
  as.factor(round(data$diabetes)),
  positive = "1",
  dnn = c("Prediction", "Reference"),
)
confusionMatrix(
  as.factor(round(measurement_model_pred)),
  as.factor(round(data$diabetes)),
  positive = "1",
  dnn = c("Prediction", "Reference"),
)
confusionMatrix(
  as.factor(round(oracle_model_pred)),
  as.factor(round(data$diabetes)),
  positive = "1",
  dnn = c("Prediction", "Reference"),
)
```




