---
title: "SylizedExample_CrimeArrests"
author: "Jonas Mikhaeil"
date: "2024-02-20"
output: html_document
---

```{r}
library("MASS")
library("rstan")
library("extraDistr")
library(ggplot2)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

## Stylized Example: Linear SEM for Arrest and Behavior

Zanger-Tishler et al. (2023) propose the following linear SEM as an
stylized model for arrests and behavior

$$
\begin{align} \nonumber    Z &\sim Normal(0,\sigma_Z) \\  \begin{bmatrix}B_0  \\B_1   \end{bmatrix} | Z  &\sim MVN  \bigg(\begin{bmatrix}\beta  Z  \\\beta Z   \end{bmatrix},\begin{bmatrix}\sigma_B^2 & \delta \\\delta & \sigma_B^2 \end{bmatrix}\bigg) \\\nonumber A_0 |Z,B_0& \sim Normal(\alpha Z + \gamma B_0, \sigma_A) \\\nonumber A_1 |Z,B_1& \sim Normal(\alpha Z + \gamma B_1, \sigma_A).\end{align} 
$$

The following code simulates data from this model:

```{r}
create_SEM_data <- function(beta, n)
{
  ### set parameters for the SEM
  alpha <- 0.4
  gamma <- 0.4
  delta <- 0.4
  ### define standard deviations. Zanger-Tishler et al. (2023) show how to choose the standard deviations of the exogenous variables such that (B0,B1,A0,A1,Z) are standardized.
  sd_z <- 1
  sd_A <- sqrt(1 - alpha ^ 2 - gamma ^ 2 - 2 * alpha * beta * gamma)
  sd_B <- sqrt(1 - beta ^ 2)
  ### define exogenous variables
  U_Z <- rnorm(n = n, mean = 0, sd = sd_z)
  U_A0 <- rnorm(n = n, mean = 0, sd = sd_A)
  U_A1 <- rnorm(n = n, mean = 0, sd = sd_A)
  sigma <- matrix(c(sd_B ^ 2, delta, delta, sd_B ^ 2), 2)
  U_B <- mvrnorm(n = n,
                 mu = rep(0, 2),
                 Sigma = sigma)
  ### calculate states
  Z <- U_Z
  B_0 <- beta * Z + U_B[, 1]
  B_1 <- beta * Z + U_B[, 2]
  A_0 = alpha * Z + gamma * B_0 + U_A0
  A_1 = alpha * Z + gamma * B_1 + U_A1
  data.frame(
    A0 = A_0,
    A1 = A_1,
    B0 = B_0,
    B1 = B_1,
    Z = Z,
    alpha = alpha,
    beta = beta,
    gamma = gamma,
    delta = delta,
    sd_z = sd_z,
    sd_A = sd_A,
    sd_B = sd_B
  )
}
```


## Hierarchichal Bayesian Measurement Model: Leakage Model


$$
\begin{align} 
    y_1 | u_1, \alpha,\gamma,\sigma_y \, &\sim \, \text{normal}(X \alpha + \gamma u_1, \sigma_y) 
    \\
     y_2 | u_2, \alpha,\gamma,\sigma_y \, &\sim \, \text{normal}(X \alpha + \gamma u_2, \sigma_y) 
     \\
      u_1 | \beta \sigma_u\, &\sim \, \text{normal}(X \beta , \sigma_u) 
      \\
      u_2 | u_1, \beta, \eta, \sigma_u\, &\sim \, \text{normal}(X \beta + \eta(u_1 - X\beta) , \sigma_u \sqrt{1-\eta^2})\\
       \sigma_y &\sim \text{normal}^+(0,1) \\
    \alpha &\sim \text{normal}(0,1) \\
    \eta &\sim \text{normal}(0,0.2) \\
    \beta &\sim \text{normal}(\beta_{true},0.1) \\
    \gamma &\sim \text{normal}(\gamma_{true},0.1).
\end{align}
$$
This measurement model is implemented in error_model_strongPriors.rstan. 
The following R code fits the model.
```{r}
data_em_sp <- list(
  N = n,
  A0 = sem_data$A0,
  A1 = sem_data$A1,
  Z = sem_data$Z,
  sd_B = sem_data$sd_B[1],
  beta = sem_data$beta[1],
  gamma = sem_data$gamma[1] + 0.1,
  sd_gamma = 0.2
)
init_fun_em_sp <-
  function() {
    list(
      B0 = rnorm(n, 0, 1),
      B1 = rnorm(n, 0, 1),
      sd_A = rhnorm(1, 1),
      eta = runif(1, 0, 1),
      alpha = rnorm(1, 0, 1),
      gamma_rand = rbeta(1, 1, 1),
      alpha_rand = rnorm(1, 0, 1)
    )
  }
fit_em_sp <-
  stan(
    file = "error_model_strongPriors.rstan",
    data = data_em_sp,
    iter = 10000,
    chains = 4,
    init = init_fun_em_sp,
    refresh = 1000
  )
RMSE_em_sp <- calculate_RMSE(fit_em_sp, sem_data)
```


## Improvement in RMSE: Plot for Figure 4

```{r}
rmse_data <- data.frame()
beta_list <-
  c(0, 0.1, 0.2, 0.25, 0.3, 0.35, 0.4, .45, 0.5, 0.55, .6)#seq(0,0.6,0.05)
for (beta in beta_list) {
  filename <- paste("data/fix/RMSE_", beta, ".Rda", sep = "")
  rmse_beta <- readRDS(filename) |> tibble::add_column(beta = beta)
  rmse_data <- rbind(rmse_data, rmse_beta)
}
```

```{r}
ggplot(data = rmse_data) + geom_line(aes(x = beta, y = RMSE_em_sp), color =
                                       "orange") +
  geom_line(aes(x = beta, y = RMSE_lm_simp), color = "#B22222") +
  geom_line(aes(x = beta, y = RMSE_lm_comp), color = "steelblue") +
  geom_line(aes(x = beta, y = RMSE_lm_oracle), color = "black") +
  scale_x_continuous(expand = c(0, 0), limits = c(-0, 0.6)) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text.x = element_text(size = 11),
    axis.title.x = element_text(size = 14),
    axis.text.y = element_text(size = 11)
  ) +
  ylab(latex2exp::TeX("RMSE")) +
  annotate(
    geom = "text",
    x = 0.1,
    y = 1.08,
    label = latex2exp::TeX("Complex"),
    color = "steelblue"
  ) +
  annotate(
    geom = "text",
    x = 0.5,
    y = 0.95,
    label = latex2exp::TeX("Simple"),
    color = "#B22222"
  ) +
  annotate(
    geom = "text",
    x = 0.1,
    y = 0.85,
    label = latex2exp::TeX("Error Model"),
    color = "orange"
  ) +
  annotate(
    geom = "text",
    x = 0.47,
    y = 0.7,
    label = latex2exp::TeX("Oracle"),
    color = "black"
  ) + xlab(latex2exp::TeX("\\beta"))
```

## Disparate Predictions: Creating plot for Figure 5

```{r}
ggplot(data = rmse_data) + geom_line(aes(x = beta, y = cor_em_sp), color =
                                       "orange") +
  geom_line(aes(x = beta, y = cor_em_sp_noZ),
            color = "orange",
            linetype = "dashed") +
  geom_line(aes(x = beta, y = cor_lm_simp), color = "#B22222") +
  geom_line(aes(x = beta, y = cor_lm_comp), color = "steelblue") +
  geom_line(aes(x = beta, y = cor_lm_oracle), color = "black") +
  scale_x_continuous(expand = c(0, 0), limits = c(-0, 0.6)) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text.x = element_text(size = 11),
    axis.title.x = element_text(size = 14),
    axis.text.y = element_text(size = 11)
  ) +
  ylab(latex2exp::TeX("$corr(Z$,Prediction Error)")) +
  annotate(
    geom = "text",
    x = 0.5,
    y = -0.2,
    label = latex2exp::TeX("Complex"),
    color = "steelblue"
  ) +
  annotate(
    geom = "text",
    x = 0.5,
    y = 0.35,
    label = latex2exp::TeX("Simple"),
    color = "#B22222"
  ) +
  annotate(
    geom = "text",
    x = 0.34,
    y = 0.05,
    label = latex2exp::TeX("Error Model with Z"),
    color = "orange"
  ) + xlab(latex2exp::TeX("\\beta")) + annotate(
    geom = "text",
    x = 0.1,
    y = -0.2,
    label = latex2exp::TeX("Error model without Z"),
    color = "orange"
  ) + xlab(latex2exp::TeX("\\beta"))
```
## Misspecification: Creating plots for Figure 6

```{r}
p_list <- c(0, 0.01, 0.1, 0.3, 0.5, 0.7, 1)
sigma_list <- c(0.01, 0.1, 0.2)
RMSE_data <- data.frame()
for (grid_index in 1:21) {
  params <- expand.grid(p = p_list, sigma_gamma = sigma_list)[grid_index, ]
  p <- params$p
  sigma_gamma <- params$sigma_gamma
  file_name <-
    paste("data/miss/miss_gamma/RMSE_",
          p,
          "_",
          sigma_gamma,
          ".Rda",
          sep = "")
  RMSE <-
    readRDS(file = file_name) |> tibble::add_column(p = p, sigma_gamma = sigma_gamma)
  RMSE_data <- rbind(RMSE_data, RMSE)
}
```

```{r}
ggplot() +
  geom_line(data = RMSE_data,
            aes(
              x = 1 + p,
              y = RMSE_em_sp,
              group = sigma_gamma,
              color = as.factor(sigma_gamma)
            )) +
  scale_colour_manual(values = c("orange", "#ff7518", "#e3ab57")) +
  #geom_line(data=RMSE_data,aes(x=p,y=RMSE_lm_simp,group=sigma_gamma,color=sigma_gamma)) +
  geom_hline(yintercept = 0.966924, color = "#B22222") +
  geom_hline(yintercept = 1.007856, color = "steelblue") +
  scale_x_continuous(
    expand = c(0, 0),
    limits = c(1, 2.2),
    breaks = c(1, 1.5, 2)
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text.x = element_text(size = 11),
    axis.title.x = element_text(size = 14),
    axis.text.y = element_text(size = 11)
  ) +
  ylab(latex2exp::TeX("RMSE")) +
  xlab(latex2exp::TeX("multiplicative misspecification of \\gamma")) + theme(legend.position = "none") +
  annotate(
    geom = "text",
    x = 1 + 0.1,
    y = 1.003,
    label = latex2exp::TeX("Complex"),
    color = "steelblue"
  ) +
  annotate(
    geom = "text",
    x = 1 + 0.1,
    y = 0.96,
    label = latex2exp::TeX("Simple"),
    color = "#B22222"
  ) +
  annotate(
    geom = "text",
    x = 1 + 0.59,
    y = 0.93,
    label = latex2exp::TeX("$\\sigma_{\\gamma} = 0.01$"),
    color = "orange"
  ) +
  annotate(
    geom = "text",
    x = 1 + 1.1,
    y = 0.957,
    label = latex2exp::TeX("$\\sigma_{\\gamma} = 0.1$"),
    color = "#ff7518"
  ) +
  annotate(
    geom = "text",
    x = 1 + 1.1,
    y = 0.943,
    label = latex2exp::TeX("$\\sigma_{\\gamma} = 0.2$"),
    color = "#e3ab57"
  )

```
```{r}
ggplot() +
  geom_line(data = RMSE_data,
            aes(
              x = 1 + p,
              y = cor_em_sp,
              group = sigma_gamma,
              color = as.factor(sigma_gamma)
            )) +
  scale_colour_manual(values = c("orange", "#ff7518", "#e3ab57")) +
  #geom_line(data=RMSE_data,aes(x=p,y=RMSE_lm_simp,group=sigma_gamma,color=sigma_gamma)) +
  geom_hline(yintercept = 0.065085590, color = "#B22222") +
  geom_hline(yintercept = -0.26287614, color = "steelblue") +
  geom_hline(yintercept = 0, color = "black") +
  
  scale_x_continuous(
    expand = c(0, 0),
    limits = c(1, 2.02),
    breaks = c(1, 1.5, 2)
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text.x = element_text(size = 11),
    axis.title.x = element_text(size = 14),
    axis.text.y = element_text(size = 11)
  ) +
  ylab(latex2exp::TeX("$corr(Z$,Prediction Error)")) + xlab(latex2exp::TeX("multiplicative misspecification of $\\gamma$")) +
  theme(legend.position = "none")
```
```{r}
p_list <- c(-1.5, -2, 1.5, 2, 0, -0.5, 0.5, -1.0, 1.0)
sigma_list <- c(0.01, 0.1, 0.2)
RMSE_data <- data.frame()
for (grid_index in 1:27) {
  params <- expand.grid(p = p_list, sigma_gamma = sigma_list)[grid_index, ]
  p <- params$p
  sigma_gamma <- params$sigma_gamma
  file_name <-
    paste("data/miss/miss_beta/RMSE_",
          p,
          "_",
          sigma_gamma,
          ".Rda",
          sep = "")
  RMSE <-
    readRDS(file = file_name) |> tibble::add_column(p = p, sigma_gamma = sigma_gamma)
  RMSE_data <- rbind(RMSE_data, RMSE)
}
```

```{r}
ggplot() +
  geom_line(data = RMSE_data,
            aes(
              x = (1 + p),
              y = RMSE_em_sp,
              group = sigma_gamma,
              color = as.factor(sigma_gamma)
            )) +
  scale_colour_manual(values = c("orange", "#ff7518", "#e3ab57")) +
  #geom_line(data=RMSE_data,aes(x=p,y=RMSE_lm_simp,group=sigma_gamma,color=sigma_gamma)) +
  geom_hline(yintercept = 0.966924, color = "#B22222") +
  geom_hline(yintercept = 1.007856, color = "steelblue") +
  scale_x_continuous(expand = c(0, 0), limits = c(-1, 3.25)) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text.x = element_text(size = 11),
    axis.title.x = element_text(size = 14),
    axis.text.y = element_text(size = 11)
  ) +
  ylab(latex2exp::TeX("RMSE")) +
  xlab(latex2exp::TeX("multiplicative misspecification of $\\beta$")) + theme(legend.position = "none") +
  annotate(
    geom = "text",
    x = 0.1,
    y = 1.003,
    label = latex2exp::TeX("Complex"),
    color = "steelblue"
  ) +
  annotate(
    geom = "text",
    x = 0.1,
    y = 0.96,
    label = latex2exp::TeX("Simple"),
    color = "#B22222"
  ) +
  annotate(
    geom = "text",
    x = 0,
    y = 0.91,
    label = latex2exp::TeX("$\\sigma_{\\gamma} = 0.01$"),
    color = "orange"
  ) +
  annotate(
    geom = "text",
    x = 2.3,
    y = 0.884,
    label = latex2exp::TeX("$\\sigma_{\\gamma} = 0.1$"),
    color = "#ff7518"
  ) +
  annotate(
    geom = "text",
    x = -0.1,
    y = 0.875,
    label = latex2exp::TeX("$\\sigma_{\\gamma} = 0.2$"),
    color = "#e3ab57"
  )
```

```{r}
ggplot() +
  geom_line(data = RMSE_data,
            aes(
              x = 1 + p,
              y = cor_em_sp,
              group = sigma_gamma,
              color = as.factor(sigma_gamma)
            )) +
  scale_colour_manual(values = c("orange", "#ff7518", "#e3ab57")) +
  #geom_line(data=RMSE_data,aes(x=p,y=RMSE_lm_simp,group=sigma_gamma,color=sigma_gamma)) +
  geom_hline(yintercept = 0.065085590, color = "#B22222") +
  geom_hline(yintercept = -0.26287614, color = "steelblue") +
  geom_hline(yintercept = 0, color = "black") +
  scale_x_continuous(
    expand = c(0, 0),
    limits = c(-1, 3.01),
    breaks = c(-1, .0, 1, 2, 3)
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text.x = element_text(size = 11),
    axis.title.x = element_text(size = 14),
    axis.text.y = element_text(size = 11)
  ) +
  ylab(latex2exp::TeX("$corr(Z$,Prediction Error)")) + xlab(latex2exp::TeX("multiplicative misspecification of $\\beta$")) + theme(legend.position = "none")
```

